Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila

### 2018-09-19 phil_paper/jmp_wjv_v2.tex
lots of good little fixes here

Figure~\ref{figure:predictedwateruse} compares the predicted histogram of consumption with observed values.  Vertical red lines indicate the price thresholds in the tariff schedule.  By roughly matching the smoothness of the distributions especially across price thresholds, this model improves on standard models in the literature, which often predict excessive bunching at thresholds (\cite{mcrae2014infrastructure}, \cite{olmstead2009reduced}, \cite{szabo2015value}).  The model has difficultly rationalizing low levels of consumption likely because households face zero prices below this threshold.\footnote{Since preference and consumption shocks are assumed to be normally distributed, the model predicts a handful of negative consumption values.  In the counterfactual exercises, I round these negative realizations to zero consumption.}  Likewise, the model also cannot account for bunching at zero (likely capturing households that go on vacation).  The model also predicts greater bunching around the upper price threshold despite little observed bunching in the data as large users tend to ignore this smaller jump in price.

Table~\ref{table:correstimates} presents correlation estimates decomposed from the standard deviation terms.  Both $\epsilon$ and $\eta$ shocks demonstrate negative correlations between neighbors for both first-movers and second-movers.  This finding suggests that neighboring households may intentionally form sharing relationships with whom their shocks are negatively correlated.  This behavior may provide an insurance benefit in the presence of non-linear prices, reducing the risk that joint consumption exceeds price thresholds each month.  Alternatively, these negative consumption findings may stem from households that frequently travel ensuring that at least one user is home to take care of the connection.  For example, given the high rate of overseas workers from the Philippines, it may be common for households to match with other households with different work schedules.  
%Since I have no information on the overseas work status of individuals, I am not able to test this hypothesis empirically.

\begin{table}
\centering
\caption{Correlation Estimates}\label{table:correstimates}
\input{tables/corr_table_group}\\
%\vspace{.3cm}
%\footnotesize{1 indexes correlations with first movers \\ while 2 indexes correlations with second movers}
\end{table}

@@

This section details how exogenous interruptions in water connections caused by leaks combine with billing records to provide a measure of hassle costs of sharing water.  The estimation strategy compares a connection's demand before disconnection to the change in demand among neighboring connections after disconnection (as a result of sharing), attributing the extent to which neighbors may not fully offset their consumption to the hassle cost, $p_h$.  Baseline preference parameters for both disconnecting connections and their neighbors are estimated in the procedure from Section~\ref{section:preferenceestimation}.  Given preference parameters estimated in Section~\ref{section:preferenceestimation}, the estimation chooses the level of hassle cost, $p_h$ that best maximizes the likelihood function given in equation (\ref{equation:hasslelikelihood}).

I now describe how the estimation sample is constructed.  In 36 months, the water provider received 106,171 complaints of excessive billing --- 34,927 of which can be precisely identified as underground pipe leaks along the service pipe through a text analysis of call transcripts.  These complaints do not appear to be spatially correlated with only a handful of leaks occurring among neighbors in the 36 months.  I consider any call identified as an excessive billing complaint as a leak.  While some calls may not explicitly refer to leaks, in all cases these calls refer to billing errors, meter malfunctions, or other surprise events that may drive disconnection.  Instances where leaks in main water pipes reduce access for entire neighborhoods at a time are coded separately and excluded from the sample.  Leaks are limited to cases with at least 3 observations before the leak in order to calculate baseline water demand for these connections.  Disconnection is defined as having 0 or missing consumption values for at least 90\% of the year following the leak date.  Neighboring connections that connect after the leak has taken place are dropped, since these connections lack a pre-period to estimate their fixed preference for water.  The estimation includes observations within a year and a half of the leak event to avoid bias from possible reconnection of these households as well as changes in the composition of neighboring households.  Since many of these households are not included in the connection survey, demographic characteristics as well as sharing behavior are imputed using averages from small areas surrounding these connections; this process introduces additional noise into the estimation.

@@

This analysis requires the assumption that the hassle cost, $p_h$, is homogenous across all households.  This assumption ensures that the hassle cost identified from the subset of leaking households can be extended to the population of households.  One concern may be that heterogenous hassle costs induce households to select into sharing relationships conditional on these costs.  Since households included in this quasi-experiment have selected into purchasing connections at baseline, they may face higher hassle costs compared to households that select into sharing.  Therefore, this exercise provides an upper-bound on hassle costs for the population.  The mobile survey finds a range of sharing access from fetching with containers to using the same plumbing, which suggests that households face different hassle costs.
The following additional assumptions are required to ensure that this empirical setting accurately captures the model described in Section~\ref{section:hasslemodel}.  These assumptions are discussed in more detail and tested empirically in Appendix~\ref{appendix:sharingcostassumptions}.
\begin{enumerate}
	\item \textit{ Connections that experience a leak are representative of the population of connections. } 
	\item \textit{ The decision to disconnect conditional on experiencing a leak is uncorrelated with water demand. } 
	\item \textit{ The timing of a leak is uncorrelated with any underlying changes in demand. } 
	\item \textit{  The leaking household's choice of a sharing partner is uncorrelated with that partner's water demand.  }	
	\item \textit{ Leaks are uncorrelated with changes in neighborhood demand. }
	\item \textit{  A leak for one household is uncorrelated with probability of a neighboring household experiencing a leak. }
\end{enumerate}

@@

Table~\ref{table:step2estimates} shows the estimation results for the hassle cost estimation.  The hassle cost is estimated to be \input{tables/mean_hassle}PhP per cubic meter of usage, which is approximately \input{tables/hassle_percentage}\unskip\% of the average tariff.  With only 548 connections that experience leaks, the estimation recovers wide standard errors for this estimate.\footnote{Since Steps 2 and 3 of the estimation rely heavily on estimated preference terms, $\gamma$, from Step 1, Appendix~\ref{appendix:incidentalparametertest} tests for bias due to noisy estimation of these incidental parameters and finds little evidence of significant bias.}  These results suggest that the time-costs or other physical challenges transporting water to share with a neighbor may play an important role in the extent to which households choose to use water from a neighbor's tap.

%, rejecting a hassle cost of zero with 10\% confidence

%One possibility may be that estimating a single average hassle cost for the population of connections masks significant heterogeneity across household density and dwelling type among other factors.\footnote{Future versions will estimate this heterogeneity directly.}

@@

\begin{table}
\centering
\caption{Predicted Hassle Cost Distribution}\label{table:step2histogram}
\includegraphics[scale=.7]{tables/hassle_cost_distribution.pdf} \\
%\vspace{cm}
\footnotesize{Excludes top and bottom 1\% outliers}
\end{table}

@@

\subsubsection{Disaggregating Connection-Level Preferences to Household-Level Preferences}

@@

With the full set of household preferences, the simulation routine first groups households into random sets of three neighbors with each of the 599 geographic regions.  The routine then orders households in the sequential game according to their utilities from using an individual connection.  This assumption captures the intuition that households who would have connected in the absence of any sharing are likely to be the first to take-up a new connection.  Random draws of the preference shocks, $\eta$ and $\epsilon$, are assigned to each household so that households can compute expected utilities over \input{tables/sim_time_periods}months for all possible water source choices.  These shocks are sampled from a multivariate normal distribution taking into account the estimated correlation between sharing partners from Table~\ref{table:correstimates}.  
%\footnote{ Future versions will not condition on observed choices in this way, and instead sample households across sharing relationships. }

@@

\begin{table}
\centering
\caption{Simulated Method of Moments Estimates}\label{table:smmestimates}
\input{tables/step_3_estimates_group}\\
\vspace{.05cm}
\footnotesize{ Standard errors are bootstrapped at the connection level.  Total Households:  \input{tables/totalhhsmm_group}  }
\end{table}

Table~\ref{table:smmestimates} presents the estimation results, finding a total estimated monthly fixed cost of \input{tables/fixedfee_group}PhP, which is around 50\% higher than the \input{tables/cfee_group}PhP monthly fixed fee paid the provider.  This finding suggests that households face sizable barriers to connecting to the water network such as maintenance land tenure insecurity, or administrative costs.  This total fixed fee is also comparable to the total variable price paid to the provider each month in Table~\ref{descriptives}, suggesting that fixed costs compose almost half of the total costs of using piped water.

The estimates also reveal a fixed cost for vendors equivalent to \input{tables/vendorfixedfee_group}PhP per month.  This likely captures fixed quality differences between vended water and other water sources.  The mean price for vended water of \input{tables/vendorprice_group}PhP per cubic meter is close to the highest tariff price in Figure~\ref{figure:tariff}.  These estimates include a sizable variance term for the vendor price, which is consistent with the large variety of small-scale water sources that are collapsed into this measure, including primarily deep wells as well as tanker trucks and water refilling stations.  These sources may also be unevenly distributed throughout the city.

Since consumer surplus from using piped water is defined relative to using from a vendor, the magnitudes of the welfare calculations may be sensitive to the parameter estimates of the vendor option, particularly the marginal price, $p_v$.  As a simple out-of-sample test, I compare this estimate of $p_v$ to the implied $p_v$ from data on total water expenditures for vendor households using additional census data from a section of Manila.\footnote{The 2011 Community Based Monitoring System data covers over 70\% of households in Pasay City --- a large area with a population of 500,000 in downtown Manila --- and includes information on income (used to compute imputed income) as well as water source and monthly expenditures.}  In this data, households using from vendors report expenditures of 365 PhP/month while households using individual connections report expenditures of 470 PhP/month and households sharing connections report expenditures of 370 PhP/month.  Assuming that the vendor expenditures only reflect the marginal prices paid for water (excluding the fixed cost), I can express average expenditures, $E_v$, by multiplying average vendor demand by the vendor price, $E_v = (\gamma - \alpha p_v) p_v$.  Plugging in expenditures, $E_v =365$ PhP/$\text{m}^{3}$, average fixed preferences for vendors, $\gamma=33$, and price sensitivity, $\alpha=.62$, as well as solving the quadratic expression, I find that this level of expenditure is consistent with vendor prices of either 15.7 PhP/$\text{m}^{3}$ or 37.53 PhP/$\text{m}^{3}$.  The latter price matches closely to the estimated price of 37.16 PhP/$\text{m}^3$, providing additional support to the structural estimates.

\subsection{Out-of-Sample Test}\label{section:oostest}

To evaluate how well the model is able to match observed substitution patterns between water sources, I investigate reduced-form evidence from a connection-fee discount program since this variation is not used explicitly in any part of the estimation.  The fundamental assumption of the model is that the decision to share a water connection balances splitting the fixed fee among many households against facing marginal hassle costs in fetching water as well as free-rider costs in splitting the water bill.  Under this assumption, reducing the fixed fee would lead to substitution away from shared connections toward individual connections.  This section uses quasi-experimental variation in the connection-fee to test this assumption.

In 2009, the government regulator issued new regulations reducing the connection fee from 6,000 to 2,400 PhP for low-income areas partially in an effort to extend access and affordability to low-income households.  Under the assumption that households reside in Manila for 80 months and face negligible credit constraints, this discount lowered the monthly fixed fee faced by households by 45 PhP on average.  In the presence of credit constraints, this estimate can be interpreted as a lower bound on the policy's impact on the monthly fixed price faced by households.  Figure~\ref{figure:discountfirststage} indicates the share of new connections receiving discounted fees.  The sudden rise by \input{tables/first_stage}percentage points coincides with the change in regulations and will be used to identify the timing of the discount policy.  A small fraction of households are observed to receive connection fee discounts prior to the program as part of other, smaller pro-poor strategies implemented by the water provider.

@@

Despite these caveats, the fact that the reduced-form and model-based estimates are similar in magnitude lends support to the counterfactual exercises in Section~\ref{section:counterfactualexercises}.  Table~\ref{table:oosestimates} calculates the impacts of this connection fee discount on water source choices as well as consumer welfare.  Consumer welfare is computed as the difference between current utility and the utility that each household would receive from using vended water.  Since utility is assumed to be quasi-linear, consumer welfare is reported in PhP/month.  Lowering the fixed fee leads to substitution away from both vendor and neighbor water sources, which improves consumer welfare especially for the bottom \input{tables/group_percentile_group}\unskip\% of users.

@@

Table~\ref{table:fixedfeediscount} summarizes the impacts of this change on source choices and consumer welfare measured in terms of the difference between current utility and utility from using the outside vendor option as in Section~\ref{section:oostest}.  The decrease in the fixed fee and increase in the tariff both act to decrease the incentives for sharing a connection and leading to a drop of \input{tables/discount_sharing_change_group}percentage points in households using water from their neighbors.  The third row indicates even greater substitution toward water vendors of around \input{tables/discount_vendor_change_group}percentage points in response to the higher tariff.  These effects combine to reduce average consumer surplus by \input{tables/discount_avg_surplus_change_group}PhP per month, which is over half of the initial fixed fee discount of \input{tables/c_fee_discount_group}PhP.  This decline translates into a \input{tables/discount_avg_welfare_change_group}percent loss of monthly household income.  Households in the bottom \input{tables/group_percentile_group}\unskip th percentile of water users also experience lower consumer surplus although the magnitude is much less, amounting to \input{tables/discount_low_welfare_change_group}percent of monthly household income.  Taken together, these results suggest that connection fee discount policies in this context can have unintended distributional consequences by reducing both access to piped water and welfare for the intended beneficiaries. %\footnote{Consumer surplus is measured relative to utility from using vended water.}  
\begin{table}
	\centering
	\caption{Fixed Fee Discount Counterfactual}\label{table:fixedfeediscount}
	\input{tables/discount_estimates_group} \\
	\footnotesize{Low Users include the bottom 50\% of water users. \\
	Surplus is measured in terms of PhP/month.}
\end{table}

@@

\begin{figure}
\centering
\caption{Optimal Tariffs Without Sharing}\label{figure:optimaltariffnosharing}
\includegraphics{tables/no_shr_tariff_groups.pdf}  
\end{figure}

@@

@@

\begin{table}
\centering
\caption{Optimal Tariffs Without Sharing}\label{table:optimaltariffnosharing}
\resizebox{\textwidth}{!}{%
\input{tables/welfare_no_sharing_groups}
}	\\
\footnotesize{$\text{ }^{*}$ The First-Best fee does not include individual transfers imposed by the regulator. \\  Social tariffs place all welfare weight on the bottom 50\% of water users. \\ Surplus is measured in terms of PhP/month.}

@@

This section calculates the optimal tariff structure in the current environment in Manila (when sharing is possible subject to the estimated hassle costs).  Figure~\ref{figure:optimaltariffsharing} plots the marginal prices associated with each counterfactual tariff.  Table~\ref{table:optimaltariffsharing} computes the impacts of these tariffs on source choices as well as welfare.  Comparing columns two and three of Table~\ref{table:optimaltariffsharing}, I find large welfare improvements on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income from shifting from the current tariff schedule to an optimal two-part tariff schedule.  These gains are mainly enjoyed by large users benefiting from a lower marginal price relative to the current tariff.  At the same time, flattening the tariff greatly reduces the implicit ``tax'' on shared connections, allowing small demand households to also benefit from lower marginal prices through their neighbors' connections.  These lower marginal prices almost compensate for the higher hassle and free-rider costs facing low-demand households, leaving surplus for these households unchanged.
%zooming in on tariffs below 20 PhP/$\text{m}^{3}$ in order to highlight distinctions between the two-part and three-part tariffs.  
\begin{figure}
\centering
\caption{Optimal Tariffs With Sharing}\label{figure:optimaltariffsharing}
\includegraphics{tables/shr_tariff_groups.pdf}
\end{figure}

@@

\begin{table}
\centering
\caption{Optimal Tariffs With Sharing}\label{table:optimaltariffsharing}
\resizebox{\textwidth}{!}{%
\input{tables/welfare_sharing_groups}
}
\footnotesize{$\text{ }^{*}$ The First-Best fee does not include individual transfers imposed by the regulator. \\  Social tariffs place all welfare weight on the bottom 50\% of water users. \\ Surplus is measured in terms of PhP/month.}
\end{table}

@@

@@

@@

>>>>

Figure~\ref{figure:predictedwateruse} compares the predicted histogram of consumption with observed values.  Vertical red lines indicate the price thresholds in the tariff schedule.  By roughly matching the smoothness of the distributions especially across price thresholds, this model improves on standard models in the literature, which often predict excessive bunching at thresholds (\cite{mcrae2014infrastructure}, \cite{olmstead2009reduced}, \cite{szabo2015value}).  The model has difficultly rationalizing low levels of consumption likely because households face zero prices below this threshold.  Since preference and consumption shocks are assumed to be normally distributed, the model predicts some negative consumption values.  In the counterfactual exercises, I round these negative realizations to zero consumption.  Likewise, the model also cannot account for bunching at zero (likely capturing households that go on vacation).  The model also predicts greater bunching around the upper price threshold despite little observed bunching in the data as large users tend to ignore this smaller jump in price.

Table~\ref{table:correstimates} presents correlation estimates decomposed from the consumption and preference shocks.  Both the consumptions shocks, $\epsilon$ and preference shocks, $\eta$ demonstrate negative correlations between neighbors for owners and buyers.  This finding suggests that neighboring households may intentionally form sharing relationships with whom their shocks are negatively correlated.  This behavior may provide an insurance benefit in the presence of non-linear prices, reducing the risk that joint consumption exceeds price thresholds each month.  Alternatively, these negative consumption findings may stem from households that frequently travel ensuring that at least one user is home to take care of the connection.  For example, given the high rate of overseas workers from the Philippines, it may be common for households to match with other households with different work schedules.  
%Since I have no information on the overseas work status of individuals, I am not able to test this hypothesis empirically.

\begin{table}
\centering
\caption{Correlation Estimates}\label{table:correstimates}
\input{tables/corr_table_group}
%\vspace{.3cm}
%\footnotesize{1 indexes correlations with first movers \\ while 2 indexes correlations with second movers}
\end{table}

@@

This section details how exogenous interruptions in water connections caused by leaks combine with billing records to provide a measure of hassle costs of sharing water.  The estimation strategy compares a connection's demand before disconnection to the change in demand among neighboring connections after disconnection (as a result of sharing), attributing the extent to which neighbors may not fully offset their consumption to the hassle cost, $H_C$.  Baseline preference parameters for both disconnecting connections and their neighbors are estimated in the procedure from Section~\ref{section:preferenceestimation}.  Given preference parameters estimated in Section~\ref{section:preferenceestimation}, the estimation chooses the level of hassle cost, $H_C$ that best maximizes the likelihood function given in equation (\ref{equation:hasslelikelihood}).

Leveraging this variation requires identifying a sample of leaking connections, the timing of leaks/disconnection, and usage for neighboring connections.  In 36 months, the water provider received 106,171 consumer complaints of excessive billing --- 34,927 of which can be precisely identified as underground pipe leaks along the service pipe through a text analysis of call transcripts.  These complaints do not appear to be spatially correlated with only a handful of leaks occurring among direct neighbors.  I consider any call identified as an excessive billing complaint as a leak.  While some calls may not explicitly refer to leaks, in all cases these calls refer to billing errors, meter malfunctions, or other surprise events that may drive disconnection.  Instances where leaks in main water pipes reduce access for entire neighborhoods at a time are coded separately and excluded from the sample.  Leaks are limited to cases with at least 3 observations before the leak in order to calculate baseline water demand for these connections.  Disconnection is defined as having 0 or missing consumption values for at least 90\% of the year following the leak date.  Neighboring connections that connect after the leak has taken place are dropped, since these connections lack a pre-period to estimate their fixed preference for water.  The estimation includes observations within a year and a half of the leak event to avoid bias from possible reconnection of these households as well as changes in the composition of neighboring households.  Since many of these households are not included in the connection survey, demographic characteristics as well as sharing behavior are imputed using averages from small areas surrounding these connections; this process introduces additional noise into the estimation.

@@

To capture heterogeneity in the hassle cost, $H_C$ is parameterized as a linear function of two summary characteristics of local neighborhoods defined as wards.  First, the share of single houses or apartment buildings in a neighborhood may increase the per-unit cost of sharing since the buildings often require longer informal pipe networks with possibly more maintenance costs as compared with duplexes.  Second, the average distance between water meters in a neighborhood provides another useful proxy for the average distance that informal pipe networks may need to extend between households.  Importantly, this distance measure does not account measure the physical distances between sharing partners or unconnected households.  These omissions are likely to overestimate the average distance between all neighbors.  Furthermore, areas with smaller physical distances are likely to increase observed sharing as well as the observed distance between meters, which would bias the estimates against finding heterogeneity in the hassle costs.  Table~\ref{table:descriptivesestsample} in Appendix~\ref{appendix:descriptivesest} includes descriptive statistics for these two measures.

One additional concern would be that households select into sharing relationships conditional on unobserved heterogeneity in sharing costs (not modeled here).  Since households included in this quasi-experiment have selected into purchasing connections at baseline, they may face higher hassle costs compared to households that select into sharing.  Therefore, this exercise may provide an upper-bound on hassle costs for the population.

%his assumption ensures that the hassle cost identified from the subset of leaking households can be extended to the population of households.  One concern may be that heterogenous hassle costs induce households to select into sharing relationships conditional on these costs.  Since households included in this quasi-experiment have selected into purchasing connections at baseline, they may face higher hassle costs compared to households that select into sharing.  Therefore, this exercise provides an upper-bound on hassle costs for the population.

%The mobile survey finds a range of sharing access from fetching with containers to using the same plumbing, which suggests that households face different hassle costs.

The following additional assumptions are required to ensure that this empirical setting accurately captures the model described in Section~\ref{section:hasslemodel}.  These assumptions are discussed in more detail and tested empirically in Appendix~\ref{appendix:sharingcostassumptions}.
\begin{enumerate}
	\item \textit{ Connections that experience a leak are representative of the population of connections. } 
	\item \textit{ The decision to disconnect conditional on experiencing a leak is uncorrelated with water demand. } 
	\item \textit{ The timing of a leak is uncorrelated with any underlying changes in demand. } 
	\item \textit{  The leaking household's choice of a sharing partner is uncorrelated with that partner's water demand.  }	
	\item \textit{ Leaks are uncorrelated with changes in neighborhood demand. }
	\item \textit{  A leak for one household is uncorrelated with probability of a neighboring household experiencing a leak. }
\end{enumerate}

@@

@@

Table~\ref{table:step2estimates} shows the coefficient estimates for the hassle cost estimation.  Consistent with the theory, the hassle cost rises with the share of households living in single homes or apartment buildings (instead of duplexes).  Similarly, increasing the average distance between households by one meter raises the predicted hassle cost by 17.59 PhP/m3.  For the average household, the hassle cost is estimated to be \input{tables/mean_hassle}PhP per cubic meter of usage, which is approximately \input{tables/hassle_percentage}\unskip\% of the average tariff.

Figure~\ref{figure:step2histogram} plots the histogram of predicted hassle costs across the population of households, finding substantial heterogeneity with a long tail of prohibitively expensive hassle costs.  Predicted hassle costs also have a correlation of \input{tables/correlation_hassle} with household fixed water preferences.\footnote{Section~\ref{section:disaggregation} discusses how connection-level fixed preferences $\gamma_C$ are disaggregated into household components, $\gamma_i$.}  This correlation has substantive implications for the impacts of counterfactual pricing policies; for example, high hassle costs may drive large users to sort away from sharing while small users can happily share water at little additional cost.

With only 548 connections that experience leaks, the estimation recovers wide standard errors for all three parameters.\footnote{Since Steps 2 and 3 of the estimation rely heavily on estimated preference terms, $\gamma$, from Step 1, Appendix~\ref{appendix:incidentalparametertest} tests for bias due to noisy estimation of these incidental parameters and finds little evidence of significant bias.}  Taken together, these results suggest that the time-costs or other physical challenges transporting water to share with a neighbor may play an important role in the extent to which households choose to use water from a neighbor's tap.

%, rejecting a hassle cost of zero with 10\% confidence
%One possibility may be that estimating a single average hassle cost for the population of connections masks significant heterogeneity across household density and dwelling type among other factors.\footnote{Future versions will estimate this heterogeneity directly.}

\begin{figure}
\centering
\caption{Predicted Hassle Cost Distribution}\label{figure:step2histogram}
\includegraphics[scale=.7]{tables/hassle_cost_distribution.pdf} \\
%\vspace{cm}
\footnotesize{Excludes top and bottom 1\% outliers}
\end{figure}

@@

\subsubsection{Disaggregating Connection-Level Preferences to Household-Level Preferences}\label{section:disaggregation}

@@

%Grouping households into sets of three produces some remainder households that I remove from the sample bringing the total number of households to \input{tables/totalhhsmm_group}.

With the full set of household preferences, the simulation routine first groups households into random sets of three neighbors within each of the 599 wards.  The routine then orders households in the sequential game according to their utilities from using an individual connection.  This assumption captures the intuition that households who would have connected in the absence of any sharing are likely to be the first to take-up a new connection.  Random draws of the preference and consumption shocks, $\eta$ and $\epsilon$, are assigned to each household so that households can compute expected utilities over \input{tables/sim_time_periods}months for all possible water source choices.  These shocks are sampled from a multivariate normal distribution taking into account the estimated correlation between sharing partners presented in Table~\ref{table:correstimates}.  
%\footnote{ Future versions will not condition on observed choices in this way, and instead sample households across sharing relationships. }

@@

\begin{table}
\centering
\caption{Simulated Method of Moments Estimates}\label{table:smmestimates}
\input{tables/step_3_estimates_group}\\
\vspace{.05cm}
\footnotesize{ Total Households:  \input{tables/totalhhsmm_group}   Total Wards:  599 \\ Standard errors are bootstrapped at the connection level.   }
\end{table}

Table~\ref{table:smmestimates} presents the estimation results, finding a total estimated monthly fixed cost of \input{tables/fixedfee_group}PhP, which is around 50\% higher than the \input{tables/cfee_group}PhP monthly fixed fee paid to the provider.  This finding suggests that households face sizable barriers to connecting to the water network such as maintenance land tenure insecurity, or administrative costs.  This total fixed fee is also comparable to the total variable price (tariff $\times$ volume) paid to the provider each month in Table~\ref{descriptives}, suggesting that fixed costs compose almost half of the total costs of using piped water.

The estimates also reveal a fixed cost for vendors equivalent to \input{tables/vendorfixedfee_group}PhP per month.  This likely captures fixed quality differences between vended water and other water sources.  The mean price for vended water of \input{tables/vendorprice_group}PhP per cubic meter is close to the highest tariff prices in Figure~\ref{figure:tariff}.  These estimates include a modest variance term for the vendor price, which is consistent with a competitive market for alternative water sources through deep wells, tanker trucks, and water refilling stations.

%These sources may also be unevenly distributed throughout the city.
%These estimates include a sizable variance term for the vendor price, which is consistent with the large variety of small-scale water sources that are collapsed into this measure, including primarily deep wells as well as tanker trucks and water refilling stations.  These sources may also be unevenly distributed throughout the city.

Since consumer surplus from using piped water is defined relative to using from a vendor, the magnitudes of the welfare calculations may be sensitive to the parameter estimates of the vendor option, particularly the marginal price, $p_v$.  As a simple out-of-sample test, Appendix~\ref{appendix:vendorpriceestimates} finds that the estimate of $p_v$ in Table~\ref{table:smmestimates} matches closely with the implied $p_v$ from data on total water expenditures for vendor households using additional census data from part of Manila.  This exercise provides additional confidence that the vendor price estimates are within a reasonable range.

%Assuming that the vendor expenditures only reflect the marginal prices paid for water (excluding the fixed cost), I can express average expenditures, $E_v$, by multiplying average vendor demand by the vendor price, $E_v = (\gamma - \alpha p_v) p_v$.  Plugging in expenditures, $E_v =365$ PhP/$\text{m}^{3}$, average fixed preferences for vendors, $\gamma=37$, and price sensitivity, $\alpha=.91$, as well as solving the quadratic expression, I find that this level of expenditure is consistent with vendor prices of either 15.7 PhP/$\text{m}^{3}$ or 37.53 PhP/$\text{m}^{3}$.  The latter price matches closely to the estimated price of 37.16 PhP/$\text{m}^3$, providing additional support to the structural estimates.

\subsection{Out-of-Sample Test: A Connection Fee Discount Program}\label{section:oostest}

A connection fee discount program implemented by the water company provides a useful setting to test the out-of-sample predictions of the economic model: do predicted water source choices match observed choices in response to a discount in the fixed fee?  According to the model, reducing the fixed fee would make individual connections more attractive relative to shared connections or water vendors.  This section compares the predicted and observed increase in connections in response to the program.

%To evaluate how well the model is able to match observed substitution patterns between water sources, I investigate reduced-form evidence from a connection-fee discount program since this variation is not used explicitly in any part of the estimation.  
%[chk] The fundamental assumption of the model is that the decision to share a water connection balances splitting the fixed fee among many households against facing marginal hassle costs in fetching water as well as free-rider costs in splitting the water bill.  Under this assumption, reducing the fixed fee would lead to substitution away from shared connections toward individual connections.  This section uses quasi-experimental variation in the connection-fee to test this assumption.

In 2009, the government regulator issued new regulations reducing the connection fee from 6,000 to 2,400 PhP for low-income areas partially in an effort to extend access and affordability to low-income households.  Under the assumption that households reside in Manila for 80 months and face negligible credit constraints, this discount lowered the monthly fixed fee faced by households by 45 PhP on average.  In the presence of credit constraints, this estimate can be interpreted as a lower bound on the policy's impact on the monthly fixed price faced by households.  Figure~\ref{figure:discountfirststage} indicates the share of new connections receiving discounted fees.  The sudden rise by \input{tables/first_stage}percentage points coincides with the change in regulations and will be used to identify the timing of the discount policy.  A small fraction of households are observed to receive connection fee discounts prior to the program as part of other, smaller pro-poor strategies implemented by the water provider.
% and these households are excluded from the main analysis

@@

Despite these caveats, the fact that the reduced-form and model-based estimates are similar in magnitude lends support to the counterfactual exercises in Section~\ref{section:counterfactualexercises}.  Table~\ref{table:oosestimates} calculates the impacts of this connection fee discount on water source choices as well as consumer welfare.  Consumer welfare is computed as the difference between current utility and the utility that each household would receive from using vended.  Since utility is assumed to be quasi-linear, consumer welfare is measured in PhP/month.  Lowering the fixed fee leads to substitution away from both vendor and neighbor water sources, which improves consumer welfare especially for the bottom \input{tables/group_percentile_group}\unskip\% of users.

@@

Table~\ref{table:fixedfeediscount} summarizes the impacts of this change on source choices and consumer welfare measured in terms of the difference between current utility and utility from using the outside vendor option as in Section~\ref{section:oostest}.  While the increase in the tariff drives large users to substitute toward vendor options, the decrease in the fixed price allows some small users to join together and purchase shared connections, resulting in no change in the share of households using from a neighbor.  The third row indicates even greater substitution toward water vendors of around \input{tables/discount_vendor_change_group}percentage points in response to the higher tariff.  These effects combine to reduce average consumer surplus by \input{tables/discount_avg_surplus_change_group}PhP per month, which is over half of the initial fixed fee discount of \input{tables/c_fee_discount_group}PhP.  This decline translates into a \input{tables/discount_avg_welfare_change_group}percent loss of monthly household income.  Households in the bottom \input{tables/group_percentile_group}\unskip th percentile of water users also experience lower consumer surplus although the magnitude is much less, amounting to \input{tables/discount_low_welfare_change_group}percent of monthly household income.  Taken together, these results suggest that connection fee discount policies in this context can have unintended distributional consequences by reducing both access to piped water and welfare for the intended beneficiaries. %\footnote{Consumer surplus is measured relative to utility from using vended water.}  
% a drop of \input{tables/discount_sharing_change_group}percentage points in households using water from their neighbors The decrease in the fixed fee and increase in the tariff both act to decrease the incentives for sharing a connection and leading to no change in households using water from their neighbors. 
\begin{table}
	\centering
	\caption{Fixed Fee Discount Counterfactual}\label{table:fixedfeediscount}
	\input{tables/discount_estimates_group} \\
	\footnotesize{Low Users include the bottom 50\% of water users. \\
	Surplus is measured in terms of PhP/month.}
\end{table}

@@

\begin{figure}[!htp]
\centering
\caption{Optimal Tariffs Without Sharing}\label{figure:optimaltariffnosharing}
\includegraphics{tables/no_shr_tariff_groups.pdf}  
\end{figure}

@@

\begin{table}[!hbp]
\centering
\caption{Optimal Tariffs Without Sharing}\label{table:optimaltariffnosharing}
\resizebox{\textwidth}{!}{%
\input{tables/welfare_no_sharing_groups}
}	\\
\footnotesize{$\text{ }^{*}$ The First-Best fee does not include individual transfers imposed by the regulator. \\  Social tariffs place all welfare weight on the bottom 50\% of water users. \\ Surplus is measured in terms of PhP/month.}
\end{table}

@@

@@

This section calculates the optimal tariff structure in the current environment in Manila (when sharing is possible subject to the estimated hassle costs).  Figure~\ref{figure:optimaltariffsharing} plots the marginal prices associated with each counterfactual tariff.  Table~\ref{table:optimaltariffsharing} computes the impacts of these tariffs on source choices as well as welfare.  Comparing columns two and three of Table~\ref{table:optimaltariffsharing}, I find large welfare improvements on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income from shifting from the current tariff schedule to an optimal two-part tariff schedule.  These gains are mainly enjoyed by large users benefiting from a lower marginal price relative to the current tariff.  At the same time, flattening the tariff greatly reduces the implicit ``tax'' on shared connections, allowing small demand households to also benefit from lower marginal prices through their neighbors' connections.  These lower marginal prices almost compensate for the higher hassle and free-rider costs facing low-demand households, leaving surplus for these households unchanged.
%zooming in on tariffs below 20 PhP/$\text{m}^{3}$ in order to highlight distinctions between the two-part and three-part tariffs.

@@

@@

\begin{figure}[!htp]
\centering
\caption{Optimal Tariffs With Sharing}\label{figure:optimaltariffsharing}
\includegraphics{tables/shr_tariff_groups.pdf}
\end{figure}

@@

\begin{table}[!hbp]
\centering
\caption{Optimal Tariffs With Sharing}\label{table:optimaltariffsharing}
\resizebox{\textwidth}{!}{%
\input{tables/welfare_sharing_groups}
}
\footnotesize{$\text{ }^{*}$ The First-Best fee does not include individual transfers imposed by the regulator. \\  Social tariffs place all welfare weight on the bottom 50\% of water users. \\ Surplus is measured in terms of PhP/month.}
\end{table}

@@

\section{Out of Sample Test for Vendor Price Estimates}\label{appendix:vendorpriceestimates}

%Since consumer surplus from using piped water is defined relative to using from a vendor, the magnitudes of the welfare calculations may be sensitive to the parameter estimates of the vendor option, particularly the marginal price, $p_v$.  As a simple out-of-sample test, I compare this estimate of $p_v$ to the implied $p_v$ from data on total water expenditures for vendor households using additional census data from a section of Manila.

As a simple out-of-sample test, I compare the estimate of $p_v$ in Table~\ref{table:smmestimates} to the implied $p_v$ from data on total water expenditures for vendor households using additional census data from a section of Manila. The 2011 Community Based Monitoring System data covers over 70\% of households in Pasay City --- a large area with a population of 500,000 in downtown Manila --- and includes information on income (used to compute imputed income) as well as water source and monthly expenditures.  In this data, households using from vendors report expenditures of 365 PhP/month while households using individual connections report expenditures of 470 PhP/month and households sharing connections report expenditures of 365 PhP/month.  I can express average marginal expenditures, $E_v$, by multiplying average vendor demand by the vendor price and adding the vendor fixed cost, $E_v = (\gamma - \alpha p_v) p_v + F_v$.  Plugging in expenditures, $E_v =365$ PhP/$\text{m}^{3}$, average fixed preferences for vendors, $\gamma=37$, and price sensitivity, $\alpha=.91$, as well as solving the quadratic expression, I find that this level of expenditure is consistent with vendor prices of either 7.9 PhP/$\text{m}^{3}$ or 32.8 PhP/$\text{m}^{3}$.  The latter price almost perfectly matches the estimated price of 32.85 PhP/$\text{m}^3$, providing additional support for the structural estimates.
